---
title: 'Intro to R: Data types, structures and manipulation with packages'
author: "Iyanu Oduwole"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
---


```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("")
library(tinytex)

```


useful link https://bookdown.org/yih_huynh/Guide-to-R-Book/other-resources.html

# Data types in R 




## Numeric

```{r}
g <- 53.5  # assign 53.3 to the variable g
g 
class(g) # class of g, data type of g
```





## Integers

```{r}
b <- 4L #Appending L to 4 
b
typeof(b)
a <- as.integer(4) 
a
typeof(a) # same as class
```



## Character

```{r}
x <- "Data Mining"
x
typeof(x)

y <- as.character(42) 
y
names <- c("John","Alex","Mo")
class(names)
```


## Logical

```{r}
a <- 3
b <- 6       
g <- a > b # is a larger than b? This returns a logical value, TRUE or FALSE
g  
class(g)
TF <- c(TRUE,FALSE,TRUE) # no qoutations, "TRUE" is a character NOT logical
class(TF)
```



## Factors

R `factor` is used to store categorical data as *levels*. It can store both character and integer types of data. These factors are created with the help of *factor()* function, by taking a vector as input.


```{r}
Education_level <- c("Below HS","HS","Some College","College Grad","Post Graduate")
class(Education_level)
Education_level <- factor(Education_level)
class(Education_level)
levels(Education_level) # levels of the vector Education_level, this is a vector of class "factor"
Education_level <- factor(Education_level, levels =  c("Below HS","HS","Some College","College Grad","Post Graduate"))
Education_level
#levels(Education_level) <- c(levels(Education_level), "Professor")

```









# Data Structures in R 

## Vector

**Vector** – A vector is a sequence of **elements** that share the same data type. These elements are known as components(elements) of a vector. A vector could be of class type `numeric`, `integer`, `logical`, `factor`, or `character`.


Vectors are defined by separating its entries by commas, and encasing them with *c()* (you can think of the command *c()* as combine or concatenate). Entries need not be numbers, but if they aren’t, you need to put the entries in quotes.


R vector comes in two parts: *Atomic vectors* and *Lists*. They have three common properties:


* *Type function* – What it is?
* *Length function* – How many elements it contains.
* *Attribute function* – Extra arbitrary metadata.



Things to keep in mind(learn)

- Creating vectors
- Accessing Elements of vectors
- Types of Vectors
- Vector Operations
- Applications of vectors



We will utilize two types of vectors in this class: *numerical* and *categorical* (nuances between nominal vs ordinal, ratio vs. interval come up infrequently). Categorical vectors have elements that represent a finite set of possible outcomes (like hair color, gender). R refers to a categorical vector as a **factor**, and stores the possible values of a factor (called **levels**) as a separate object.



### Indexing Vectors

There are four forms of vector indexing used to extract a subset of vectors:

1. Positive Integers Vector – Indicates selected elements.
2. Negative Integers Vector – Indicates rejected elements.
3. Character Strings Vector – Used only for vectors with named elements.
4. Logical Values Vector – They are the result of evaluated conditions.


```{r}
vec <- c(5,10,2,3,1,11,NA)  # c for concatenate, NA is for missing value
vec[c(1,5,6,9)] # elements correponding to 1,5,6,9 indices
vec[2:5] # elements corresponding to indices 2:5
vec[-3] # all but the third element
vec[vec<5] # all elements that are less than 5
```



### The `which` command

You may only want to work with specific elements of a vector (like only the ones
larger than 5). The `which` command returns the positions within the vector of
elements that meet a certain condition (not the values of the elements themselves).


```{r}
x <- c(1,8,2,-10,-4,5,5,1,6,9,-2,2)
which(x==2)#positions where elements are equal to 2
which(x<0)#positions where elements are less than 0
x[ which(x<0) ]#elements where elements are less than 0
mean(x[which(x >0)])#average of all elements in x larger than 0
```





### Logical conditions

* `equals` ==
* `not equals` !=
* `greater than` >
* `greater than or equal to` >=
* `less than` <
* `less than or equal to` <=
* is missing `is.na()`, e.g., `which(is.na(x))` tells you the positions of missing data
* is not missing `!is.na()`, e.g., `which(!is.na(x))` tells you the positions of data that exists
* belongs to a list of values `%in%`



```{r}
5==5
which(x==5) # returns position of values in x that are equal to 5
```


**You can combine conditions:**

* "and" `&`
* "or" `|` (shift backslash key)


```{r}
x <- c(1,8,2,-10,-4,NA,5,NA,6,9,-2,NA)#redefine x and put some missing values in it
which( x >=0 & x <= 4)#positions where x is between 0 and 4, inclusive
which( x <=0 | x == 2)#positions where x is 0 or less, or 2
which( x <=0 | x == 2|is.na(x))#positions where x is 0 or less, equal to 2, or missing
```




### The `%in%` shortcut

```{r}
x <- c(1,5,8,2,-10,-4,NA,5,NA,6,9,-2,NA)#redefine x and put some missing values in it
which( x %in% c(1,2,5,12))#positions where x is 1, 2, 5, or 12
x[ which( x %in% c(1,2,5,12))]#values of x that are either 1, 2, 5, or 12
x[ which( x %in% seq(0,100,by=2))]#values of x that are even numbers between 0 and 100
x[ which( x <= 5)]#values of x that are at most 5
x[ which( x %in% (1:10)^2)]#values of x that are square numbers between 1 and 100
```





### Summarizing vectors

```{r}
vec <- c(5,10,2,3,1,11,NA) 
summary(vec)  #5 number summary
length(vec) #number of elements
length( unique(vec) ) #number of non-duplicated values
min(vec)  #returns NA because NAs exist in the data
min(vec,na.rm=TRUE) #need na.rm=TRUE when NAs are in data to get the values out
mean(vec)
mean(vec,na.rm=TRUE) # remove the NAs and return the mean
```





## Sequences and Patterns in R

Sequences of numbers can be made in R with the `seq` command. This takes 3 of 4 possible arguments `from`, `to`, `by`, and `length`.


```{r}
seq(from=10,to=11,length=5)
```


The `rep(x,times)` command is useful for creating a vector with a certain pattern, e.g., creating a pattern `x repeated times` times.


```{r}
rep(0,25)
```




## factor vectors


```{r}
x <- factor( c("dog","dog","cat","cat","cat","dog","dog") )  #a factor
levels(x)  #two possible values are cat and dog
## ----Adding a level------------------------------------------------------
x[8] <- "fish"  #cannot add fish to the end of this vector since that is not a valid level

#lets start over
x <- factor( c("dog","dog","cat","cat","cat","dog","dog") )  #a factor
levels(x) <- c( levels(x), "fish")  #use the levels command to append the list of valid levels
summary(x)
x[8] <- "fish" #success, because "fish" is level that exists
summary(x)

## Renaming levels of a factor
x <- factor(c("a","a","b","b","b","","","c","d"))
x  #the blank is for all intents and purposes invisible
levels(x)  #Notice that a blank is shown as back to back quotation marks 
levels(x)[1] <- "missing"   #Renaming the blank level (1st in the list) to be missing
x
levels(x)

## Renaming levels of a factor safe
x <- factor(c("a","a","b","b","b","","","c","d"))
levels(x)[ which( levels(x) == "" )] <- "missing"   #Safe way of renaming the blank level to be missing
x
levels(x)

## Dealing with blanks
x <- factor(c("a","a","b","b","b","","","c","d"," "," ",""," "))
x  #the blank is for all intents and purposes invisible
levels(x)  #Notice that a blank is shown as back to back quotation marks 
#following code is general and replaces blanks, single space, double space, and tab
levels(x)[ which( levels(x) %in% c(""," ","  ","\t")) ] <- "missing"   
x
levels(x)

## Combining levels dangerous
x <- factor(c("a","b","b","b","b","c","d","e","e"))
levels(x)
table(x)
#Imagine we want to combine a, c, and d (which appear once) into a level called other
levels(x)[c(1,3,4)] <- "other"   #a is the 1st, c is the 3rd, d is the 4th levels of x
table(x)

## Combining levels safe
x <- factor(c("a","b","b","b","b","c","d","e","e"))
#Imagine we want to combine a, c, and d (which appear once) into a level called other
#Use which, %in%, and give a vector of levels you want to combine
levels(x)[ which(levels(x) %in% c("a","c","d")) ] <- "other"   
table(x)
```




If x is a factor, you can get a frequency table, examine its levels, make a bar chart, etc. See

* `table(x)` - a list of levels and how many times they appear
* `table(x)/length(x)` - a relative frequency table (percentages)
* `levels(x)` - the possible values of the factor
* `plot(x)` - bar chart
* `barplot(table(x))` - bar chart









### Note on selecting elements of factors


For factors, elements are text expressions. When selecting elements using `==` or
`%in%`, quotes need to surround whatever is on the right of == or %in%. Otherwise,
`R` will assume that what is on the right is a vector, and use the vector’s contents.

```{r}
x <- factor( c("f","u","n","k","y","b","u","t","t","e","r","s","i","p","s"))
which(x == "u") #things on right of == need to be in quotes
which(x %in% c("b","a","s")) #things on right of == need to be in quotes
s <- c("p","e","t","r","i","e")
which(x %in% s)#s not in quotes, so R uses what makes up the vector
```



### Alternative: the subset command


The function which determines the locations of certain elements in a vector. The result is
usually encased in to extract those elements. E.g., `x[which(x>5)]` gives the elements
that are bigger than 5. The function subset conveniently combines these two steps.

```{r,eval=FALSE}
subset(name of vector, condition regarding elements of vector)
```


```{r}
x <- c(1,8,2,-10,-4,NA,5,NA,6,9,-2,NA)#redefine x and put some missing values in it
subset(x,x>5) #reports the values of x that are bigger than 5
subset(x,!is.na(x))#reports the values of x that are not missing
x <- factor( c("a","b","b","b","a","a","cat","cat","dog") )
subset(x,x=="a" | x=="b")
subset(x,x %in% letters) #letters is a vector a, b, c, d, etc.,
subset(x,x=="hi")#reports elements that equal hi, of which there are none
```










## Matrices

**Matrices** – A matrix is basically a two-dimensional array of numbers or other types of mathematical objects. We can do operations such as addition and multiplication on the matrix in R. 

```{r}
mat <- matrix(1:25,nrow = 5, ncol = 5) # create a 5 by 5 matrix with the elements 1:25
mat
```


```{r}
X <- c(16.92, 24.03, 7.61, 15.49, 11.77)    
Y <- c(8.37, 12.93, 16.65, 12.2, 13.12)
comb <- cbind(X,Y) # column bind
colnames(comb) # column names
rownames(comb) <- LETTERS[1:5] # row names, LETTERS is built in vector
comb
```


### Indexing Matrices
```{r}
print(comb)
comb[1,2]
```


Creating R matrix?


```{r}
arrang_row <- matrix(c(4:15), nrow = 4, byrow = TRUE) #Creating our matrix and arranging it by row
arrang_row #Printing our arranged matrix
arrang_col <- matrix(c(4:15), nrow = 4, byrow = FALSE)       #Creating our matrix and arranging it by column
arrang_col    #Printing our arranged matrix 
```



Row names and column names for a matrix
```{r}
rows <- c("row1", "row2", "row3", "row4")     #Creating our character vector of row names
cols <- c("col1", "col2", "col3")          #Creating our character vector of column names
mat <- matrix(c(4:15), nrow = 4, byrow = TRUE, dimnames = list(rows, cols) ) 
mat
```



**Modifying Elements of a Matrix**

```{r}
mat    #Displaying values of matrix mat
mat[2,3] <- 20  #Assigning value 20 to the element at 2nd row and 3rd column
mat       #Displaying our modified matrix.

mat[mat == 4] <- 0     #Replacing elements that are equal to 4 with 0
mat        #Displaying our modified matrix ‘mat’

mat[mat < 10] <- 0       #Replacing elements that are less with 10 with 0
mat        #Displaying modified matrix ‘mat’
```



**Addition of Rows and Columns**
```{r}
new_mat <- matrix(1:12, nrow = 3, ncol = 3)
new_mat
```

Now, we will add a column to our matrix ‘new_mat’ using *cbind()* function as follows:
```{r}
cbind(new_mat, c(1,2,3)) # cbind for column bind
```

We can also add a row using the *rbind()* function as follows:

```{r}
rbind(new_mat, c(1,2,3)) # rbind for row bind
```



### R Matrix Operations


```{r}
mat1 <- matrix(data = 1:8, nrow = 4, ncol = 4) #Creating our first matrix mat1 
mat2 <- matrix(data = 1:16, nrow = 4, ncol = 4) #Creating our second matrix mat2
sum <- mat1 + mat2 #Adding our two matrices
sum #Printing the sum
sub <- mat1 - mat2 #Adding our two matrices
sub #Printing the sum
prod <- mat1*4 #Multiplying matrix mat1 with constant value 4
prod #Printing the product
prod <- mat1*mat2 #Multiplying matrix mat1 with mat2 - this is element-wise multiplication
prod #Printing the product
div <- mat1/mat2 #Division of mat1 and mat2 
div #Printing the division
```






## Lists  

R list is the object which contains elements of different types – like *strings*, *numbers*, *vectors* and *another list* inside it. R list can also contain a matrix or a function as its elements. The list is created using the *list()* function in R. In other words, a list is a generic vector containing other objects.


```{r}
vec <- c(1,2,3)
char_vec <- c("R", "Python", "Matlab", "SAS")
logic_vec <- c(TRUE, FALSE, TRUE, FALSE)
our_list <- list(vec, char_vec, logic_vec)
our_list
```


How do we create a *list*? Let’s create a list containing *string, numbers, vectors* and *logical values*.

```{r}
list_data <- list("Red", "White", c(1,2,3), TRUE, 22.4)
print(list_data)
```


Naming elements of a *list*.

```{r}
data_list <- list(c("Jan","Feb","Mar"), matrix(c(1,2,3,4,-1,9), nrow = 2),list("Red",12.3))
names(data_list) <- c("Month", "Matrix", "Misc")
print(data_list)
```



How to Access R List Elements? 



```{r}
data_list[1] # first element of the list
data_list$Month
```

Altering elements in a list?


```{r}
data_list[4] <- "New element" # adding a fourth element to the list
data_list[4] <- NULL # remove the fourth element
data_list[3] <- "updated element" # update the third element
```






## Arrays

We will skip this. Not very often used in R












## Data-frames


A data frame is a two-dimensional array. Unlike a matrix, the data we store in the columns of the data frame can be of various types. It means one column might be a numeric variable, another might be a factor, and a third might be a character variable. All columns have to be of the same length.


* The column names should be non-empty.
* The row names should be unique.
* The data frame can hold the data which can be a numeric, character or of factor type.
* Each column should contain the same number of data items.



### R Data Frame Operations

Creating a data frame using *data.frame()*.

```{r}
employee_data <- data.frame(
  employee_id = c (1:5),
  employee_name = c("James","Harry","Shinji","Jim","Oliver"),
  sal = c(642.3,535.2,681.0,739.0,925.26),
  join_date = as.Date(c("2013-02-04", "2017-06-21", "2012-11-14", "2018-05-19","2016-03-25")),
  stringsAsFactors = FALSE)
employee_data
str(employee_data) # Get the Structure of the R Data Frame
```


dimensions of a data frame:
```{r}
dim(employee_data)
```


**Indexing a data frame** : we essentially index data frames just like we would an R matrix.



```{r}
employee_data[1:2,] # rows 1 and 2
```


Expanding a data frame


```{r}
employee_data$dept <- c("IT","Finance","Operations","HR","Administration")
employee_data

# add a data frame with a new employee
new_employee_data <- data.frame(
employee_id = c (6),
employee_name = c("Emma"),
sal = c(523.0),
join_date = as.Date(c("2015-06-22")),
dept = "Sales",
stringsAsFactors = FALSE
)

new_employee_data

updated_employee_data <- rbind(employee_data,new_employee_data) # row binding the two data frames
updated_employee_data
summary(updated_employee_data)
```



**Existing data frames in R**: There are many data sets in R housed in different packages.

```{r}
head(iris) # first 6 rows of the iris dataset
str(iris) # structure of the iris dataset
class(iris) # class of the iris dataset
```



**Subsetting a data frame**: This is a very important concept to understand. You will often be interested in a subset of a dataset, be it subset of rows or subset of columns. The *subset()* function does exactly that. The first argument of the *subset()* function is the data frame.

```{r}
levels(iris$Species) # the three types of Species (Levels of Species variable)
```

How can we subset to just the data for Species=="setosa"? Lets use the *subset* function


```{r}
iris_setosa <- subset(iris,Species=="setosa") # first argument is the data frame, followed by the condition
head(iris_setosa)
subset(iris,Species=="setosa" & Petal.Length > 1.7) # Setosa species with Petal.Length > 1.7
subset(iris,Species=="setosa" & Petal.Length > 1.7, select = c("Sepal.Length","Sepal.Width","Petal.Length","Petal.Width"))
```

**Data manipulations  can be made effective using packages**


### What is Tidyverse?

- Tidyverse is a collection of R packages designed for data science.
- It provides a consistent and coherent set of functions for data manipulation, visualization, and modeling.
- Tidyverse emphasizes tidy data principles, making data cleaning and analysis easier and more intuitive.

---

# Load the tidyverse
library(tidyverse)
search()
##  [1] ".GlobalEnv"        "package:forcats"   "package:stringr"  
##  [4] "package:dplyr"     "package:purrr"     "package:readr"    
##  [7] "package:tidyr"     "package:tibble"    "package:ggplot2"  
## [10] "package:tidyverse" "package:stats"     "package:graphics" 
## [13] "package:grDevices" "package:utils"     "package:datasets" 
## [16] "package:methods"   "Autoloads"         "package:base"
### Key Tidyverse Packages

- **dplyr**: For data manipulation
- **ggplot2**: For data visualization
- **tidyr**: For data tidying
- **purrr**: For functional programming
- **readr**: For data import
- **tibble**: For modern data frames

---

### Using Tidyverse Functions

- Tidyverse functions follow a consistent syntax and approach.
- The pipe operator `%>%` is used to chain together multiple operations.
- Functions are designed to work seamlessly with each other, allowing for a streamlined workflow.
Tutorial adopted from: "https://jcoliver.github.io/learn-r/012-intro-tidyverse.html"

Here we are using the iris data were collected by botanist Edgar Anderson and used in the early statistical work of R.A. Fisher.

Our goal : Explore the data stored in the pre-loaded iris dataset


```{r, warning=FALSE}
#install.packages("tidyverse")
library(tidyverse)
head(iris)
glimpse(iris)
head(iris)

```

**Summarizing the data**
So what will we need? Break this down into the component parts.
Means
Standard errors
For each species
For each trait


```{r}
# The hard way by getting means for a single column
setosa_mean <- mean(iris$Sepal.Length[iris$Species == "setosa"])
versicolor_mean <- mean(iris$Sepal.Length[iris$Species == "versicolor"])
virginica_mean <- mean(iris$Sepal.Length[iris$Species == "virginica"])

# Add these back into a data.frame
iris_means <- data.frame(Species = c("setosa", "versicolor", "virginica"),
                         SL_mean = c(setosa_mean, versicolor_mean, virginica_mean))

# There is a better way with tidy verse
# Group the iris data by values in the Species column
iris_grouped <- group_by(iris, Species)
head(iris_grouped)
iris_means <- summarise(iris_grouped, SL_mean = mean(Sepal.Length))
# Really cool part is that you can pipe these together
iris_means <- iris %>%
  group_by(Species) %>%
  summarize(SL_mean = mean(Sepal.Length))
glimpse(iris_means)
# if you want to summarise the standard error as well
iris_means <- iris %>%
  group_by(Species) %>%
  summarize(SL_mean = mean(Sepal.Length),
            SL_se = sd(Sepal.Length)/sqrt(n()))
iris_means
```

**Reorganize the data so each row only has data for a single trait**
```{r}
#This is known as “long” format, where each row only has a single trait observation. To make this data conversion, we use the the pivot_longer function:
iris_long <- pivot_longer(data = iris, 
                    cols = -Species,
                    names_to = "trait",
                    values_to = "measurement")

# compute that with the Iris means
iris_means <- iris %>%
  pivot_longer(cols = -Species, 
               names_to = "trait", 
               values_to = "measurement") %>%
  group_by(Species, trait) %>%
  summarize(trait_mean = mean(measurement),
            trait_se = sd(measurement)/sqrt(n()))

iris_means
# joining data frame

# Assume we have two copies of the Iris dataset
iris1 <- iris
iris2 <- iris

# Add a new column to iris2   and iris1
iris1$ID <- sample(1:150, 150)
iris2$ID <- sample(1:150, 150)

# Perform inner join
inner_join_result <- inner_join(iris1, iris2, by = "ID")

# Perform outer join
outer_join_result <- full_join(iris1, iris2, by = "ID")
# perform merge
merge_result <- merge(iris1, iris2, by = "ID", all = TRUE)

#Note : inner_join() performs an inner join on the common column "ID", and full_join() performs an outer join, resulting in a dataset containing all rows from both iris1 and iris2. merge() performs a merge operation on the common column "ID", and the all = TRUE argument specifies that it should perform a full outer join, resulting in a dataset containing all rows from both iris1 and iris2

```

**loading and manipulating data with tidyverse part 2**
```{r, results='hide'}
surveys <- read.csv("~/Documents/R/colloqium_presentation/surveys.csv")
glimpse(surveys)
# f we want to create a new object with this smaller version of the data

surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

# To create a new column of weight in kg with mutate
surveys %>%
  mutate(weight_kg = weight / 1000)
glimpse(surveys)
surveys %>%
  mutate(weight_kg = weight / 1000,
         weight_kg2 = weight_kg * 2)

# The first few rows of the output are full of NAs, so if we wanted to remove those we could insert a filter() in the chain:

surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) 

glimpse(surveys)

surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))
```

**Conditional statements with Data manipulation**
```{r}
# ifelse statements 
surveys$hindfoot_cat <- ifelse(surveys$hindfoot_length < 29.29, "small", "big")
head(surveys$hindfoot_cat)
# you can also use mutate with case_when
surveys %>% 
  mutate(hindfoot_cat = case_when(
    hindfoot_length > 29.29 ~ "big",
    TRUE ~ "small"
  )) %>% 
  select(hindfoot_length, hindfoot_cat) %>% 
  head()
glimpse(surveys)
# To adjust for the NAs
surveys %>% 
  mutate(hindfoot_cat = case_when(
    hindfoot_length > 29.29 ~ "big",
    is.na(hindfoot_length) ~ NA_character_,
    TRUE ~ "small"
  )) %>% 
  select(hindfoot_length, hindfoot_cat) %>% 
  head()

```


# We will go back to for and while loops if we have time
## for loop 

This is what a for loop looks like.

```{r,eval=FALSE}
for(looping.variable in vector.of.values) {
command 1
command 2
command 3
etc.
}
```


```{r}
words <- c("hi","there")
for(p in words) {
info <- c( p, nchar(p) ) # info contains the word and its number of characters
print(info)
}
```


Let us use a for loop to redefine the elements in positions 2, 3, 7 of a vector x.
Specifically, the element in position i should equal 2*i+5.
```{r}
x <- c(2.1,9.1,5.9,4.8,3.7,1.6,3.1,3.8,2.9,6.9,4.9,8.9,7.1) # Give x an initial definition
for(i in c(2,3,7)) {
x[i] <- 2*i + 5
}
x
```

Imagine I have 10 students in a class and I need to calculate a weighted average of
their homework. It would be tedious to do one by one (a lot of copy/pasting), so a
loop will help out.
```{r}
#First, make up a gradebook with random numbers between 0.7-1
set.seed(474)
GRADES <- matrix( runif(50,.7,1),nrow=10,ncol=5 )#10 students, 5 assignments
wg <- rep(0,nrow(GRADES))#Initialize weighted grades to be a vector of ten 0s (one for each student)
for (i in 1:length(wg)) {
wg[i] <- 0.5*mean(GRADES[i,1:4])+0.5*GRADES[i,5]
}
wg
```

**Note**: **the initialization step is important**. Running x[2] <- 9 will fail if x
does not already exist in the global environment. To put something inside a vector,
R needs to be aware of that vector’s existence first!



**Another example:** In forecasting, a popular method for “smoothing out" a jumpy time series is to take a moving average of (say) the two observations before and after the current one.
Imagine we have a time series called demands that contains the demands of a
product over 50 weeks. We’d like to create series of moving averages.

```{r}
#Make up a bogus time series of 50 demands by randomly picking numbers 5-20
set.seed(577); demands <- sample(5:20,size=50, replace=TRUE)
moving.avg <- rep(NA,50)#Initialize moving.avg to be 50 NAs
for (time in 3:48) {
moving.avg[time] <- mean(demands[(time-2):(time+2)] )
}
moving.avg #Some elements are NA by design
```






**We can also do nested for loops**
```{r}
mat <- matrix(1:10, 2) # nrow=2, byrow=FALSE
mat
for (id1 in seq(nrow(mat))) { # loop through the rows
  for (id2 in seq(ncol(mat))) { # loop through the columns
    print(mat[id1, id2])
  }
}
```


## while loop 


In a *for()* loop, you give the code a specific list of values to loop over (1, 2, 3, . . . ,
to the number of rows in the dataframe, all elements of a vector of words, etc.).
Sometimes you do not know how many times the loop will need to execute, but
you’ll know when it’s done. "While such and such is true, continue to do this and
that".

This is the job of a *while()* loop. A while loop allows you to repeat a set of
commands as long as some logical condition is TRUE.


Examples:

* Betting until you are broke or double your money. Perhaps you start with $10,
bet $1 each time, and you want to keep track of the total number of games
you get to play (and perhaps the money you had after each bet). The
probability of winning is 18/38 and losing is 20/38. Recall that sample(
c(-1,1),size=1,prob=c(20/38,18/38)) will randomly generate a -1 or a
1 with the desired probabilities.


```{r,eval=FALSE}
commands before loop
while( logical condition ) {
  Repeat this chunk of code
}
commands after loop
```



**Example:** Vegas Roulette problem. In English: “while the amount of money is larger than 0,
randomly determine the outcome of the next bet and adjust the amount of money
accordingly and increment the total number of bets placed by 1".

```{r}
money <- 10 # vector that will keep track of bankroll
number.of.bets <- 0 # Initialize a variable to keep track of the number of bets
set.seed(577) # set random number seed for reproducibility
while(min(money)>0){
  number.of.bets <- number.of.bets + 1 #Increment the number of bets that have been made
  result <- sample( c(-1,1),size=1,prob=c(20/38,18/38))#get random outcome of bet
  money <- c(money,tail(money,1) + result) #add an additional element and update the current amount
}
number.of.bets
money
```



Since money is a vector whose $i$-th element is the amount of money we have going
into bet $i$, we can make a plot to see the time evolution of the money we have.

```{r}
money <- 10; number.of.bets <- 0; set.seed(21)
while ( min(money) > 0) {
    number.of.bets <- number.of.bets + 1
    result <- sample( c(-1,1),size=1,prob=c(20/38,18/38))
    money <- c(money,tail(money,1) + result) 
}
number.of.bets
bet.number <- 1:length(money)
plot(money ~ bet.number ,xlab="Bet Number",ylab="Money",type="l")
```







